---
title: "Reading PRIMER PERMANOVA outputs"
format: html
vignette: >
  %\VignetteIndexEntry{Reading PRIMER PERMANOVA outputs}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(primertools)
```

## Overview

This vignette demonstrates how to extract key results from a PRIMER
PERMANOVA text output file using `primertools`.

Specifically, we show how to read:

- the *PERMANOVA table of results*
- the *estimates of components of variation*

## Example PERMANOVA file

The package ships with a small example PERMANOVA output file.

```{r}
f <- system.file("extdata", "PERMANOVA1.txt", package = "primertools")

cat(readLines(f), sep = "\n")
```

## PERMANOVA table of results

```{r}
tab <- read_permanova_results_table(f)
tab
```

## Components of variation

```{r}
covs <- read_permanova_components_of_variation(f)
covs
```

## Notes

- These functions are designed for PRIMER v7–v8 text outputs.
- They ignore auxiliary sections such as EMS tables and pseudo-F constructions.
- The word "Unique" is added to the `Unique perms` column manually (because the "Unique" is dropped when reading the table).
