---
title: "Reading PRIMER SIMPER outputs"
format: html
vignette: >
  %\VignetteIndexEntry{Reading PRIMER SIMPER outputs}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(primertools)
```

## Overview

This vignette shows how to read a PRIMER SIMPER text output file into a structured list
of within-group and between-group contribution tables.

It also demonstrates helper functions that convert the nested list output into tidy
tibbles that are easier to filter, join, tabulate, and plot.

## Read an example file shipped with the package

```{r}
simper_path <- system.file("extdata", "SIMPER1.txt", package = "primertools")
simper_path

x <- read_simper(simper_path)

names(x)
names(x$within)
names(x$between)
```

## Within-group contributions

The `within` element is a named list of data frames, one per group.

```{r}
x$within$A
```

## Between-group contributions

The `between` element is a named list of data frames, one per group contrast.

```{r}
x$between$`A vs H`
```

## Helper: convert within-group results to a single tibble

`tidy_simper_within()` stacks the within-group tables into one tibble and adds a `group`
column. This makes it easy to filter by cumulative contribution and to work with the
results using standard tidyverse workflows.

```{r}
d_within <- tidy_simper_within(x, cum_max = 90)

d_within
```

Example: show the top 10 taxa per group by contribution.

```{r}
d_within |>
  dplyr::group_by(group) |>
  dplyr::slice_max(contrib, n = 10, with_ties = FALSE) |>
  dplyr::ungroup()
```

## Helper: convert between-group results to a single tibble

`tidy_simper_between()` stacks the between-group tables into one tibble and adds a
`contrast` column. It also computes `delta_abund = av_abund_1 - av_abund_2` to indicate
which group has the higher average abundance for each taxon.

```{r}
d_between <- tidy_simper_between(x, cum_max = 70)

d_between
```

Example: extract the top contributors for one contrast.

```{r}
d_between |>
  dplyr::filter(contrast == "A vs H") |>
  dplyr::arrange(desc(contrib)) |>
  dplyr::slice_head(n = 15)
```

## Helper: make within-facet ordering for plots

When plotting SIMPER contributions in faceted panels (e.g., one panel per group),
ggplot2 orders factor levels globally. `add_taxon_group_order()` adds a plotting-only
factor (`taxon_group`) that is ordered within each group.

```{r}
d_within_plot <- d_within |>
  add_taxon_group_order()

d_within_plot |>
  dplyr::select(taxon, group, contrib, taxon_group) |>
  dplyr::slice_head(n = 10)
```

### Example plot: within-group contributions

```{r}
library(ggplot2)

ggplot(d_within_plot, aes(x = contrib, y = taxon_group)) +
  geom_col() +
  facet_wrap(~ group, scales = "free_y") +
  tidytext::scale_y_reordered() +
  labs(
    title = "Within-group SIMPER",
    subtitle = "Taxa contributing to within-group Bray–Curtis similarity (cumulative cutoff applied)",
    x = "Contribution to within-group similarity (%)",
    y = NULL
  )
```

## Notes on interpretation

SIMPER is a descriptive decomposition of average similarity/dissimilarity into taxon
contributions. Contributions are not independent and SIMPER does not provide measures
of statistical uncertainty.
